Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-05-28 advances/informe2.tex
Up informe 2

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that economic
institutions influence people's normative attitudes
(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
find that countries with stronger public provision or more expansive
welfare states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

@@

Although redistribution typically refers to the state's capacity to
reallocate resources from the advantaged to the vulnerable, market
institutions also play a central role in distributing socially valuable
goods and rewards (Koos \& Sachweh, 2019; Lindh \& McCall, 2023). As
Polanyi (1957) observed, economic integration in capitalist societies is
primarily organized through market exchange, governed by a
self-regulating price system embedded in institutional frameworks. These
institutions are not mere aggregates of individual behavior, but social
realities endowed with rules, mechanisms, and normative meanings that
shape everyday thinking (Koos \& Sachweh, 2019; Immergut, 1998). In this
sense, the economic order is mirrored in a moral economy: collectively
shared norms and beliefs about justice in distribution, embedded and
reinforced through institutions (Mau, 2015; Svallfors, 2006). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(Lindh \& McCall, 2023). In many countries, privatization and
commodification have expanded market logic into core areas of social
reproduction, such as healthcare, education, pensions, and social
security, deepening inequality in access to these services (Gingrich,
2011; Ferre, 2023). Yet public support for market-based welfare
provision has grown, even in traditional welfare states, particularly
among higher-income groups who view private alternatives as more
efficient or higher in quality (Busemeyer \& Iversen, 2020). This shift
calls for broader inquiry into how institutions like markets structure
access to resources and legitimize inequalities (Lindh \& McCall, 2023;
Mau, 2015).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also examines the beliefs individuals hold about the causes
of social inequality, the normative principles that underpin those
beliefs, the factors that shape them, and their consequences for
attitudes and behavior (Kluegel \& Smith, 1981). In this regard,
empirical research on distributive justice focuses on individuals'
conceptions of how goods and rewards \emph{should} be distributed in
society (Jasso et al., año). This line of inquiry allows for examining
the extent to which economic inequality is perceived as just, or in a
certain sense, legitimate (Castillo, 2011). One prominent strand
investigates the legitimacy of wage inequality, particularly salary
differences across occupations (Jasso, 1978; Castillo, 2011; Wegener,
1999). Another examines how individuals justify market-generated
inequalities in access to core social services such as healthcare,
education, and pensions. Here, legitimacy stems from the belief that
access to these goods should follow market-based criteria (Castillo et
al., 2024; Lindh, 2015). In such views, these services are treated as
legitimate commodities; goods that can be traded, priced, and evaluated
through market logic (Busemeyer \& Iversen, 2020).

Market justice preferences refer to normative beliefs that legitimize
the idea that access to core social services---such as healthcare,
education, or pensions---should be determined by market-based criteria.
Following Janmaat's (2013, p.~359) distinction, these preferences fall
under the category of ``beliefs,'' understood as normative ideas about
what inequality should look like, as opposed to ``perceptions,'' which
refer to subjective evaluations of existing inequality. Market justice
preferences reflect the view that access to these services should depend
on individuals' ability to pay, thus justifying inequalities generated
by market mechanisms (Kluegel \& Smith, 1999; Lindh, 2015). The concept
draws on Lane's (1968) classic contrast between market justice and
political justice: the former is grounded in the idea of earned
deserts---where rewards reflect effort, productivity, and skill---while
the latter prioritizes need and equality, typically expressed in welfare
state policies. Lane argued that markets and states differ in purpose
(efficiency vs.~need), logic (individual vs.~collective), and fairness
criteria (merit vs.~equality). Market justice assumes that markets are
neutral, self-regulating systems in which fair procedures yield outcomes
proportional to merit. Inequality, from this perspective, is not only
expected but legitimate---as long as it arises from fair competition. In
this way, market justice offers a moral lens through which individuals
can view the commodification and stratified access to social services as
fair and justified (Kluegel \& Smith, 1999; Lindh, 2015).

The empirical study of market justice preferences has relied on diverse
strategies to operationalize how individuals evaluate inequalities
produced by market allocation. One common approach assesses whether
people consider it fair that access to key social services---such as
healthcare, education, or pensions---depends on income. This strategy
builds on foundational work by Kluegel and Smith (1981), who explored
the normative justifications behind support for economic inequality.
More recent studies have expanded this logic beyond income to include
market-mediated access to welfare goods. For instance, Immergut and
Schneider (2020) and von dem Knesebeck et al.~(2016) examine whether
respondents believe it is just that those with higher incomes receive
better healthcare. Similarly, Lee and Stacey (2023) apply this approach
to educational opportunities. These studies typically use targeted
survey items that ask respondents to evaluate specific distributive
scenarios, allowing for comparative analyses across countries or welfare
regimes. Complementing these efforts, more recent research has
introduced composite indicators designed to capture broader orientations
toward market-based allocation. For example, Castillo et al.~(2024)
propose a single-item measure that summarizes individual support for
income-based access across multiple domains---health, education, and
pensions---within the Chilean context. Together, these instruments aim
to capture to what extent individuals perceive market-generated
inequality as justified.

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles (Svallfors, 2007; Koos \&
Sachweh, 2019; Lindh, 2015). For example, Lindh (2015) finds that
individuals from the service class are more likely to support
market-based access to healthcare and education than skilled and
unskilled workers across 17 relatively affluent countries. In a
comparative analysis, Svallfors (2007) observes that this expected class
pattern appears clearly only in Sweden, where support for private
education and healthcare varies systematically by class. Busemeyer
(2014) similarly shows that support for private education is stronger
among high-income groups, while Immergut and Schneider (2020) and von
dem Knesebeck et al.~(2016) report comparable findings for healthcare,
suggesting that wealthier individuals perceive private provision as a
means of maintaining relative advantage. In Chile, Otero and Mendoza
(2024) show that individuals with higher income and university education
express stronger support for market allocation in healthcare, education,
and pensions. Beliefs about inequality and political orientation also
matter. In Chile, Castillo et al.~(2024) show that individuals who have
strong meritocratic beliefs are more likely to support market-based
distribution, while Lee and Stacy (2023) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

However, individual characteristics alone do not fully explain variation
in support for market justice. A growing body of cross-national research
highlights the importance of institutional arrangements in shaping these
preferences. For instance, Immergut and Schneider (2020) find that in
countries with higher public spending on healthcare, individuals are
less likely to view income-based access as fair. Similarly, Busemeyer
(2014) shows that increased public investment in education is associated
with lower support for privatized provision. Conversely, Lindh (2015)
finds that in countries with more market-oriented welfare systems,
support for market-based distribution tends to be higher, suggesting
that individual attitudes often align with institutional outputs. These
findings are consistent with neo-institutionalist and policy feedback
theories (Campbell, 2012), which argue that institutions do more than
redistribute resources---they also shape the normative categories
through which individuals assess who is deserving of support (Immergut,
1990). In this view, dominant values and preferences are both shaped by
and embedded in institutional configurations (Busemeyer, 2014),
reinforcing the notion that institutions are not neutral structures, but
active producers of the moral frameworks that legitimize or challenge
inequality.

@@

The study of social mobility, its drivers and consequences has long been
central to sociology. Classic theorists explored not only movement
across social hierarchies but also its broader implications for class
conflict, norm stability, and institutional change (Breen \& Ermisch,
2024). Sorokin (1927) introduced the concept formally, defining mobility
as the shift of individuals, values, or objects between positions within
a stratification system, and distinguishing between horizontal and
vertical forms. Later work differentiated intergenerational from
intragenerational mobility, as well as absolute mobility---driven by
structural change---from relative mobility, which captures the extent to
which origins constrain destinations (Eyles et al., 2022). In Latin
America, research shows high absolute but low relative mobility (Bucca,
2016): although educational expansion and economic modernization have
enabled some upward movement, status reproduction remains strong,
especially among elites (Torche, 2015; López-Roldán \& Fachelli, 2021).
This reflects deep structural inequality, segmented education systems,
stratified labor markets, and legacies of dependent development that
restrict access to mobility channels and reinforce the intergenerational
transmission of advantage.

Beyond mapping mobility patterns, growing research has examined its
subjective and attitudinal effects. Mobility effects---defined as
outcomes resulting from movement between origin and destination classes
(Breen \& Ermisch, 2024)---have long attracted theoretical interest.
Sorokin's (1959) dissociative hypothesis posits that mobility, whether
upward or downward, may produce psychological strain due to conflicting
norms between class contexts, leading to lower life satisfaction.
Similarly, Lenski (1954) argued that status inconsistency---mismatches
among education, income, and occupation---can undermine well-being.
These ideas underpin extensive empirical work linking intergenerational
mobility to outcomes such as life satisfaction, mental health, and
stress (Hadjar, Gugushvili, Präg \& Gugushvili).

Research on intergenerational social mobility has increasingly examined
its effects on attitudes toward economic inequality. A key area of
inquiry focuses on how upward and downward mobility influence support
for redistribution, though findings remain mixed. Alesina et al.~(2018)
show that individuals with pessimistic expectations about their
mobility---particularly those anticipating downward movement---are more
likely to support generous redistributive policies. Similarly, Ares
(2020) finds that upwardly mobile individuals tend to be less supportive
of state-led redistribution compared to those who have experienced
downward mobility. Comparative studies by Schmidt (2011) and Gugushvili
(2016) likewise report that subjective upward mobility is associated
with weaker preferences for redistribution, while downward mobility
strengthens redistributive support. However, recent causal evidence from
Breen and Ermisch (2024) suggests the opposite: upward mobility may
increase redistributive preferences, while downward mobility may reduce
them.

Beyond redistribution, other studies have explored the impact of
mobility on broader beliefs about inequality. Gugushvili (2016) finds
that upwardly mobile individuals are more likely to adopt
individualistic attributions of poverty and to legitimize income
inequality, particularly in post-socialist societies. Similarly, Bucca
(2016) shows that subjective upward mobility reinforces individualistic
explanations of wealth in seven Latin American countries. At the macro
level, Shariff et al.~(2016) demonstrate that higher levels of national
economic mobility correlate with greater tolerance of inequality. In
contrast, Day and Fiske (2019) find that low perceived mobility
undermines belief in meritocracy and a just world, thereby weakening
system justification. Taken together, this body of research suggests
that mobility shapes attitudes toward inequality and justice through
multiple, and sometimes contradictory, mechanisms.

The literature on the effects of social mobility has proposed various
mechanisms to explain how and why changes in social position may
influence individual outcomes (Freyr \& Rhem, año). One of the most
prominent is the self-interest mechanism, which posits that individuals
who experience upward or downward mobility undergo a shift in their
material interests, thereby altering their perceptions and preferences
(Ares, 2020; Langsæther, Evans \& O'Grady, 2021; Freyr \& Rhem, año).
Closely related to this logic is the Prospect of Upward Mobility (POUM)
hypothesis, which suggests that individuals may oppose redistribution
not because of their current position, but because they anticipate
improving their status in the future (REF). A second line of explanation
draws on the framework of theories of socialization. Within this
tradition, hypotheses such as acculturation, socialization, and status
maximization propose that individuals adjust their attitudes based on
the norms and values of either their class of origin or their
destination, or a combination of both (Jaime-Castillo \&
Marqués-Perales, 2019). However, most of these mechanisms focus
primarily on the indirect effects of origin and destination positions,
rather than on the direct effect of experience of movement itself. In
response to this gap, recent research has highlighted the role of
self-serving bias in causal attribution processes, suggesting that
individuals tend to explain their mobility trajectories in ways that
justify their current position, which in turn shapes their beliefs and
preferences (Schmidt, 2011; Gugushvili, 2016; Molina, Bucca \& Macy,
2019).

The self-serving bias mechanism builds on a value-oriented perspective,
emphasizing that individuals' experiences of social mobility shape their
causal attributions, which in turn influence their beliefs about justice
and distributive preferences (Gugushvili, 2014). Causal attribution
refers to the process through which individuals generate explanations
for their own behavior and outcomes, as well as those of others
(Gugushvili, year). In this view, people's interpretations of economic
inequality depend on whether they believe such disparities reflect
unequal individual contributions. Individuals who adopt an internal
attribution framework tend to see success or failure as rooted in
personal characteristics such as effort, talent, or merit. In contrast,
those who regard inequality as unjust are more likely to adopt an
external attribution model, viewing outcomes as the result of structural
barriers beyond individual control (Kluegel \& Smith, 1999). This
mechanism, often described as intrapersonal causal attribution, reflects
how people explain their own socioeconomic positions---typically
attributing their successes to internal qualities while blaming failures
on external circumstances (Miller \& Ross, 1975). Over time, individuals
may revise their beliefs and attitudes: while early views are shaped by
their social origin, these are later adjusted in light of personal
experiences of mobility and the perceived role of ascribed versus
achieved factors in determining socioeconomic outcomes (Gugushvili,
year).

Empirical research has provided support for the self-serving bias
mechanism in shaping redistributive preferences and attitudes toward the
legitimacy of inequality. For instance, Schmidt (2011) finds that
individuals who experience upward mobility are more likely to interpret
their success as the result of personal effort or merit, and
consequently perceive redistribution as less necessary. Conversely,
individuals who experience downward mobility tend to attribute their
decline to external circumstances---such as structural inequality or
unemployment---and show stronger support for redistribution. In a
comparative analysis across different welfare domains, Gugushvili
demonstrates that upward mobility is associated with lower support for
government spending on housing and pensions, while individuals who
experience downward mobility express lower support for healthcare and
education spending, but favor increased investment in housing and
pensions, reflecting the material nature of these domains. Moreover,
Gugushvili (2016) finds that upward mobility is linked to greater
justification of income inequality, suggesting that improved social
standing reinforces an attributional view in which success is seen as
the result of individual characteristics---thus legitimizing inequality
as a fair outcome.

@@

@@

>>>>

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that economic
institutions influence people's normative attitudes
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}), studies find
that countries with stronger public provision or more expansive welfare
states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

@@

Although redistribution typically refers to the state's capacity to
reallocate resources from the advantaged to the vulnerable, market
institutions also play a central role in distributing socially valuable
goods and rewards (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
2019}; \citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As
Polanyi (\citeproc{ref-polanyi_great_1975}{1975}) observed, economic
integration in capitalist societies is primarily organized through
market exchange, governed by a self-regulating price system embedded in
institutional frameworks. These institutions are not mere aggregates of
individual behavior, but social realities endowed with rules,
mechanisms, and normative meanings that shape everyday thinking
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: collectively shared
norms and beliefs about justice in distribution, embedded and reinforced
through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In many
countries, privatization and commodification have expanded market logic
into core areas of social reproduction, such as healthcare, education,
pensions, and social security, deepening inequality in access to these
services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
for market-based welfare provision has grown, even in traditional
welfare states, particularly among higher-income groups who view private
alternatives as more efficient or higher in quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
This shift calls for broader inquiry into how institutions like markets
structure access to resources and legitimize inequalities
(\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also examines the beliefs individuals hold about the causes
of social inequality, the normative principles that underpin those
beliefs, the factors that shape them, and their consequences for
attitudes and behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \&
Smith, 1981}). In this regard, empirical research on distributive
justice focuses on individuals' conceptions of how goods and rewards
\emph{should} be distributed in society
(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
2016}). This line of inquiry allows for examining the extent to which
economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
One prominent strand investigates the legitimacy of wage inequality,
particularly salary differences across occupations
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Another
examines how individuals justify market-generated inequalities in access
to core social services such as healthcare, education, and pensions.
Here, legitimacy stems from the belief that access to these goods should
follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

Market justice preferences refer to normative beliefs that legitimize
the idea that access to core social services---such as healthcare,
education, or pensions---should be determined by market-based criteria.
Following Janmaat's (\citeproc{ref-janmaat_subjective_2013}{2013, p.
359}) distinction, these preferences fall under the category of
``beliefs,'' understood as normative ideas about what inequality should
look like, as opposed to ``perceptions,'' which refer to subjective
evaluations of existing inequality. Market justice preferences reflect
the view that access to these services should depend on individuals'
ability to pay, thus justifying inequalities generated by market
mechanisms (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al.,
1999}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The concept draws
on Lane's (\citeproc{ref-lane_market_1986}{1986}) classic contrast
between market justice and political justice: the former is grounded in
the idea of earned deserts---where rewards reflect effort, productivity,
and skill---while the latter prioritizes need and equality, typically
expressed in welfare state policies. Lane argued that markets and states
differ in purpose (efficiency vs.~need), logic (individual
vs.~collective), and fairness criteria (merit vs.~equality). Market
justice assumes that markets are neutral, self-regulating systems in
which fair procedures yield outcomes proportional to merit. Inequality,
from this perspective, is not only expected but legitimate---as long as
it arises from fair competition. In this way, market justice offers a
moral lens through which individuals can view the commodification and
stratified access to social services as fair and justified
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

The empirical study of market justice preferences has relied on diverse
strategies to operationalize how individuals evaluate inequalities
produced by market allocation. One common approach assesses whether
people consider it fair that access to key social services---such as
healthcare, education, or pensions---depends on income. This strategy
builds on foundational work by Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}), who explored the normative
justifications behind support for economic inequality. More recent
studies have expanded this logic beyond income to include
market-mediated access to welfare goods. For instance, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) examine whether
respondents believe it is just that those with higher incomes receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply this approach to
educational opportunities. These studies typically use targeted survey
items that ask respondents to evaluate specific distributive scenarios,
allowing for comparative analyses across countries or welfare regimes.
Complementing these efforts, more recent research has introduced
composite indicators designed to capture broader orientations toward
market-based allocation. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) propose a single-item
measure that summarizes individual support for income-based access
across multiple domains---health, education, and pensions---within the
Chilean context. Together, these instruments aim to capture to what
extent individuals perceive market-generated inequality as justified.

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
from the service class are more likely to support market-based access to
healthcare and education than skilled and unskilled workers across 17
relatively affluent countries. In a comparative analysis, Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
expected class pattern appears clearly only in Sweden, where support for
private education and healthcare varies systematically by class.
Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
that support for private education is stronger among high-income groups,
while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
report comparable findings for healthcare, suggesting that wealthier
individuals perceive private provision as a means of maintaining
relative advantage. In Chile, Otero and Mendoza
(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
higher income and university education express stronger support for
market allocation in healthcare, education, and pensions. Beliefs about
inequality and political orientation also matter. In Chile, Castillo et
al. (\citeproc{ref-castillo_socialization_2024}{2024}) show that
individuals who have strong meritocratic beliefs are more likely to
support market-based distribution, while Lee and Stacy
(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

However, individual characteristics alone do not fully explain variation
in support for market justice. A growing body of cross-national research
highlights the importance of institutional arrangements in shaping these
preferences. For instance, Immergut and Schneider
(\citeproc{ref-immergut_it_2020}{2020}) find that in countries with
higher public spending on healthcare, individuals are less likely to
view income-based access as fair. Similarly, Busemeyer
(\citeproc{ref-busemeyer_skills_2014}{2014}) shows that increased public
investment in education is associated with lower support for privatized
provision. Conversely, Lindh (\citeproc{ref-lindh_public_2015}{2015})
finds that in countries with more market-oriented welfare systems,
support for market-based distribution tends to be higher, suggesting
that individual attitudes often align with institutional outputs. These
findings are consistent with neo-institutionalist and policy feedback
theories (\citeproc{ref-campbell_institutional_2020}{Campbell, 2020}),
which argue that institutions do more than redistribute resources---they
also shape the normative categories through which individuals assess who
is deserving of support
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In this
view, dominant values and preferences are both shaped by and embedded in
institutional configurations
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014}), reinforcing the
notion that institutions are not neutral structures, but active
producers of the moral frameworks that legitimize or challenge
inequality.

@@

The study of social mobility, its drivers and consequences has long been
central to sociology. Classic theorists explored not only movement
across social hierarchies but also its broader implications for class
conflict, norm stability, and institutional change
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}). Sorokin
(\citeproc{ref-sorokin_social_1927}{1927}) introduced the concept
formally, defining mobility as the shift of individuals, values, or
objects between positions within a stratification system, and
distinguishing between horizontal and vertical forms. Later work
differentiated intergenerational from intragenerational mobility, as
well as absolute mobility---driven by structural change---from relative
mobility, which captures the extent to which origins constrain
destinations (\citeproc{ref-eyles_social_2022}{Eyles et al., 2022}). In
Latin America, research shows high absolute but low relative mobility
(\citeproc{ref-bucca_merit_2016}{Bucca, 2016}): although educational
expansion and economic modernization have enabled some upward movement,
status reproduction remains strong, especially among elites
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). This
reflects deep structural inequality, segmented education systems,
stratified labor markets, and legacies of dependent development that
restrict access to mobility channels and reinforce the intergenerational
transmission of advantage.

Beyond mapping mobility patterns, growing research has examined its
subjective and attitudinal effects. Mobility effects---defined as
outcomes resulting from movement between origin and destination classes
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024})---have long
attracted theoretical interest. Sorokin's (1959) dissociative hypothesis
posits that mobility, whether upward or downward, may produce
psychological strain due to conflicting norms between class contexts,
leading to lower life satisfaction. Similarly, Lenski
(\citeproc{ref-lenski_status_1954}{1954}) argued that status
inconsistency---mismatches among education, income, and occupation---can
undermine well-being. These ideas underpin extensive empirical work
linking intergenerational mobility to outcomes such as life
satisfaction, mental health, and stress
(\citeproc{ref-gugushvili_heterogeneous_2024}{Gugushvili, 2024};
\citeproc{ref-hadjar_does_2015}{Hadjar \& Samuel, 2015};
\citeproc{ref-prag_subjective_2021}{Präg \& Gugushvili, 2021}).

Research on intergenerational social mobility has increasingly examined
its effects on attitudes toward economic inequality. A key area of
inquiry focuses on how upward and downward mobility influence support
for redistribution, though findings remain mixed. Alesina et al.
(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
individuals with pessimistic expectations about their
mobility---particularly those anticipating downward movement---are more
likely to support generous redistributive policies. Similarly, Ares
(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
individuals tend to be less supportive of state-led redistribution
compared to those who have experienced downward mobility. Comparative
studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017})
likewise report that subjective upward mobility is associated with
weaker preferences for redistribution, while downward mobility
strengthens redistributive support. However, recent causal evidence from
Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the
opposite: upward mobility may increase redistributive preferences, while
downward mobility may reduce them.

Beyond redistribution, other studies have explored the impact of
mobility on broader beliefs about inequality. Gugushvili
(\citeproc{ref-gugushvili_intergenerational_2016}{2016b}) finds that
upwardly mobile individuals are more likely to adopt individualistic
attributions of poverty and to legitimize income inequality,
particularly in post-socialist societies. Similarly, Bucca
(\citeproc{ref-bucca_merit_2016}{2016}) shows that subjective upward
mobility reinforces individualistic explanations of wealth in seven
Latin American countries. At the macro level, Shariff et al.
(\citeproc{ref-shariff_income_2016}{2016}) demonstrate that higher
levels of national economic mobility correlate with greater tolerance of
inequality. In contrast, Day and Fiske
(\citeproc{ref-day_movin_2017}{2017}) find that low perceived mobility
undermines belief in meritocracy and a just world, thereby weakening
system justification. Taken together, this body of research suggests
that mobility shapes attitudes toward inequality and justice through
multiple, and sometimes contradictory, mechanisms.

The literature on the effects of social mobility has proposed various
mechanisms to explain how and why changes in social position may
influence individual outcomes
(\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025}). One of the
most prominent is the self-interest mechanism, which posits that
individuals who experience upward or downward mobility undergo a shift
in their material interests, thereby altering their perceptions and
preferences (\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
Closely related to this logic is the Prospect of Upward Mobility (POUM)
hypothesis, which suggests that individuals may oppose redistribution
not because of their current position, but because they anticipate
improving their status in the future
(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}). A second line
of explanation draws on the framework of theories of socialization.
Within this tradition, hypotheses such as acculturation, socialization,
and status maximization propose that individuals adjust their attitudes
based on the norms and values of either their class of origin or their
destination, or a combination of both
(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). However, most of these mechanisms focus
primarily on the indirect effects of origin and destination positions,
rather than on the direct effect of experience of movement itself. In
response to this gap, recent research has highlighted the role of
self-serving bias in causal attribution processes, suggesting that
individuals tend to explain their mobility trajectories in ways that
justify their current position, which in turn shapes their beliefs and
preferences
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-molina_its_2019}{Molina et al., 2019};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}).

The self-serving bias mechanism builds on a value-oriented perspective,
emphasizing that individuals' experiences of social mobility shape their
causal attributions, which in turn influence their beliefs about justice
and distributive preferences
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Causal
attribution refers to the process through which individuals generate
explanations for their own behavior and outcomes, as well as those of
others (\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
2016b}). In this view, people's interpretations of economic inequality
depend on whether they believe such disparities reflect unequal
individual contributions. Individuals who adopt an internal attribution
framework tend to see success or failure as rooted in personal
characteristics such as effort, talent, or merit. In contrast, those who
regard inequality as unjust are more likely to adopt an external
attribution model, viewing outcomes as the result of structural barriers
beyond individual control (\citeproc{ref-kluegel_beliefs_1981}{Kluegel
\& Smith, 1981}). This mechanism, often described as intrapersonal
causal attribution, reflects how people explain their own socioeconomic
positions---typically attributing their successes to internal qualities
while blaming failures on external circumstances
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Over
time, individuals may revise their beliefs and attitudes: while early
views are shaped by their social origin, these are later adjusted in
light of personal experiences of mobility and the perceived role of
ascribed versus achieved factors in determining socioeconomic outcomes
(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}).

Empirical research has provided support for the self-serving bias
mechanism in shaping redistributive preferences and attitudes toward the
legitimacy of inequality. For instance, Schmidt
(\citeproc{ref-schmidt_experience_2011}{2011}) finds that individuals
who experience upward mobility are more likely to interpret their
success as the result of personal effort or merit, and consequently
perceive redistribution as less necessary. Conversely, individuals who
experience downward mobility tend to attribute their decline to external
circumstances---such as structural inequality or unemployment---and show
stronger support for redistribution. In a comparative analysis across
different welfare domains, Gugushvili
(\citeproc{ref-gugushvili_subjective_2017}{2017}) demonstrates that
upward mobility is associated with lower support for government spending
on housing and pensions, while individuals who experience downward
mobility express lower support for healthcare and education spending,
but favor increased investment in housing and pensions, reflecting the
material nature of these domains. Moreover, Gugushvili
(\citeproc{ref-gugushvili_intergenerational_2016c}{2016a}) finds that
upward mobility is linked to greater justification of income inequality,
suggesting that improved social standing reinforces an attributional
view in which success is seen as the result of individual
characteristics---thus legitimizing inequality as a fair outcome.

@@

\bibitem[\citeproctext]{ref-ares_changing_2020}
Ares, M. (2020). Changing classes, changing preferences: How social
class mobility affects economic preferences. \emph{West European
Politics}, \emph{43}(6), 1211--1237.
\url{https://doi.org/10.1080/01402382.2019.1644575}

@@

\bibitem[\citeproctext]{ref-sorokin_social_1927}
Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
